---
title: "Benchmarks"
description: "Real, measured timings on this machine: filtering a file bigger than you want to load, round-tripping a table to disk, and the honest cost of the database round trip on small data."
---

```{r}
#| label: setup
#| include: false
source("_setup.R")
suppressPackageStartupMessages(library(microbenchmark))
has_arrow <- requireNamespace("arrow", quietly = TRUE)
has_fst   <- requireNamespace("fst", quietly = TRUE)
has_qs    <- requireNamespace("qs", quietly = TRUE)
knitr::opts_knit$set(root.dir = tempdir())
```

Every other page in this site uses small inputs so it renders quickly. This one doesn't. Every number below is measured on this machine, right now, by the code shown -- not picked to flatter the package. Where `dbdt` loses, that's reported too, along with why.

```{r}
#| label: versions
cat("data.table.ext:", as.character(utils::packageVersion("data.table.ext")), "\n")
cat("data.table:    ", as.character(utils::packageVersion("data.table")), "\n")
cat("duckdb:        ", as.character(utils::packageVersion("duckdb")), "\n")
cat("R:             ", R.version.string, "\n")
cat("cores:         ", parallel::detectCores(), "\n")
```

```{r}
#| label: results-accumulator
#| include: false
results <- list()
record <- function(section, op, baseline, baseline_s, other_s) {
  results[[length(results) + 1]] <<- data.frame(
    section = section, operation = op, baseline = baseline,
    baseline_s = unname(baseline_s), dbdt_s = unname(other_s),
    speedup = unname(baseline_s) / unname(other_s)
  )
}
```

## A file bigger than you want to load {#file-scan}

The scenario this package is built for: someone hands you a file export, and you need one number out of it. `dbdt_csv()`/`dbdt_parquet()` register the file as a lazy view -- DuckDB reads only what a query touches -- versus the usual "load the whole thing, then filter." Format matters here more than it might seem, so both are tested rather than just the more flattering one.

```{r}
#| label: make-files
set.seed(1)
n <- 5e6
pos <- data.table(
  transaction_id = seq_len(n),
  store_id       = sample(1:200, n, replace = TRUE),
  region         = sample(c("APAC", "EMEA", "AMER", "LATAM"), n, replace = TRUE),
  sku            = sprintf("SKU-%05d", sample(1:5000, n, replace = TRUE)),
  qty            = sample(1:5, n, replace = TRUE),
  unit_price     = round(runif(n, 2, 200), 2)
)
fwrite(pos, "pos.csv")
if (has_arrow) arrow::write_parquet(pos, "pos.parquet")
cat(sprintf(
  "%s rows -- %.1f MB as CSV%s\n",
  format(n, big.mark = ",", scientific = FALSE), file.size("pos.csv") / 1024^2,
  if (has_arrow) sprintf(", %.1f MB as Parquet", file.size("pos.parquet") / 1024^2) else ""
))
```

The task, against both files: total revenue for one store, in one region -- a needle in a haystack (1 store of 200, in 1 region of 4). Each variant is timed three times; the median is reported, since a cold first read is noisier than the steady state.

```{r}
#| label: timing-helper
rep_time <- function(f, times = 3) {
  ts <- vapply(seq_len(times), function(i) { gc(FALSE); system.time(f())[["elapsed"]] }, numeric(1))
  stats::median(ts)
}
```

### Parquet: pruning is real {#parquet}

Parquet stores data column-by-column, with per-row-group statistics, so a reader *can* skip most of the file for a selective filter on a couple of columns. `dbdt_parquet()` gets that for free from DuckDB; the fair baseline is `arrow::read_parquet()` also asked to select only the needed columns, not a naive full read.

```{r}
#| label: bench-parquet
#| eval: !expr has_arrow
t_arrow <- rep_time(function() {
  full <- as.data.table(arrow::read_parquet("pos.parquet", col_select = c("store_id", "region", "qty", "unit_price")))
  ans_arrow <<- full[store_id == 42 & region == "APAC", .(revenue = sum(qty * unit_price))]
})
t_dbdt_pq <- rep_time(function() {
  v <- dbdt_parquet("pos.parquet")
  ans_dbdt_pq <<- v[store_id == 42 & region == "APAC", .(revenue = sum(qty * unit_price))]
})
cat(sprintf("arrow::read_parquet(col_select = ...) + filter: %.3fs (median of 3)\n", t_arrow))
cat(sprintf("dbdt_parquet() + filter:                        %.3fs (median of 3)\n", t_dbdt_pq))
record("File scan (Parquet)", "filter + aggregate (5M rows)", "arrow::read_parquet(col_select) + filter", t_arrow, t_dbdt_pq)
```

`r if (has_arrow) sprintf("A real %.1fx here", t_arrow / t_dbdt_pq) else "Skipped (arrow not installed)"` -- DuckDB pushes both the column selection *and* the `store_id`/`region` filter into the Parquet reader itself, so it decodes a fraction of a fraction of the file. `arrow::read_parquet()` decodes the four requested columns for every row before `data.table` gets to filter any of them; the win comes from never decoding the 4.99M rows the filter throws away, not from a faster codec.

### CSV: the same trick, no advantage {#csv}

CSV is plain text, row by row, with no stored statistics and no column-oriented layout -- there is nothing analogous to prune. Every engine has to scan and parse every row to find out which ones match, and `fread()`'s parallel parser is a hard baseline to beat at that job.

```{r}
#| label: bench-csv
t_fread <- rep_time(function() {
  full <- fread("pos.csv")
  ans_fread <<- full[store_id == 42 & region == "APAC", .(revenue = sum(qty * unit_price))]
})
loaded_mb <- as.numeric(object.size(fread("pos.csv"))) / 1024^2
t_dbdt_csv <- rep_time(function() {
  v <- dbdt_csv("pos.csv")
  ans_dbdt_csv <<- v[store_id == 42 & region == "APAC", .(revenue = sum(qty * unit_price))]
})
cat(sprintf("fread() + filter:      %.3fs (median of 3, materializes ~%.0f MB in R)\n", t_fread, loaded_mb))
cat(sprintf("dbdt_csv() + filter:   %.3fs (median of 3, materializes nothing until the answer)\n", t_dbdt_csv))
record("File scan (CSV)", "filter + aggregate (5M rows)", "fread() + filter", t_fread, t_dbdt_csv)
```

```{r}
#| label: correctness-check
#| include: false
stopifnot(isTRUE(all.equal(ans_fread$revenue, ans_dbdt_csv$revenue)))
if (has_arrow) stopifnot(isTRUE(all.equal(ans_arrow$revenue, ans_dbdt_pq$revenue)))
```

`r if (t_dbdt_csv <= t_fread) "dbdt_csv() actually came out ahead here too" else sprintf("fread() wins this one -- %.1fx", t_dbdt_csv / t_fread)` -- an honest result, not the one this package would pick if it could. `dbdt_csv()` still never held the full table in R (`fread()`'s copy above is real memory, not a hypothetical), which matters once the file stops comfortably fitting -- but on a file this size, on this machine, that headroom didn't come with a speed advantage. Reach for `dbdt_csv()` for a file too big to load at all, or when the source already writes Parquet; reach for `fread()` for a CSV that fits and one query.

```{r}
#| label: cleanup-files
#| include: false
file.remove("pos.csv")
if (has_arrow) file.remove("pos.parquet")
```

## Round-tripping a table to disk {#io}

The other common case: saving a table so a later session can pick it up ([Connect & load](duckdt-connect.qmd#persist)), against the usual ways to do that in R. Ten string columns of mixed cardinality, since text is where formats diverge most -- an all-numeric table would mostly measure memcpy speed.

```{r}
#| label: io-data
rand_strings <- function(n, len) Reduce(paste0, replicate(len, sample(letters, n, replace = TRUE), simplify = FALSE))
make_data <- function(n) {
  set.seed(1)
  data.table(
    id           = sprintf("ID%010d", seq_len(n)),
    category     = sample(sprintf("cat_%02d", 1:20), n, replace = TRUE),
    region       = sample(c("APAC", "EMEA", "AMER", "LATAM"), n, replace = TRUE),
    status       = sample(c("active", "inactive", "pending", "closed"), n, replace = TRUE),
    country_code = sample(sprintf("C%03d", 1:50), n, replace = TRUE),
    product_code = rand_strings(n, 6),
    tag          = sample(sprintf("tag_%03d", 1:100), n, replace = TRUE),
    email_domain = sample(c("gmail.com", "outlook.com", "yahoo.com", "company.com", "example.org"), n, replace = TRUE),
    description  = rand_strings(n, 12),
    notes        = rand_strings(n, 20)
  )
}
n_io <- 2e6
dt_io <- make_data(n_io)
```

```{r}
#| label: io-bench
time_it <- function(f) { gc(FALSE); t0 <- proc.time()[["elapsed"]]; f(); proc.time()[["elapsed"]] - t0 }
io_rows <- list()
add_io <- function(format, op, secs, size_bytes) {
  io_rows[[length(io_rows) + 1]] <<- data.frame(
    format = format, op = op, seconds = round(secs, 3),
    rows_per_sec = round(n_io / secs), file_mb = round(size_bytes / 1024^2, 1)
  )
}

add_io("saveRDS/readRDS", "write", time_it(function() saveRDS(dt_io, "b.rds")), file.size("b.rds"))
add_io("saveRDS/readRDS", "read",  time_it(function() readRDS("b.rds")), file.size("b.rds"))

add_io("fwrite/fread", "write", time_it(function() fwrite(dt_io, "b.csv")), file.size("b.csv"))
add_io("fwrite/fread", "read",  time_it(function() fread("b.csv")), file.size("b.csv"))

if (has_arrow) {
  add_io("arrow (parquet)", "write", time_it(function() arrow::write_parquet(dt_io, "b.parquet")), file.size("b.parquet"))
  add_io("arrow (parquet)", "read",  time_it(function() as.data.table(arrow::read_parquet("b.parquet"))), file.size("b.parquet"))
}
if (has_fst) {
  add_io("fst", "write", time_it(function() fst::write_fst(dt_io, "b.fst")), file.size("b.fst"))
  add_io("fst", "read",  time_it(function() fst::read_fst("b.fst", as.data.table = TRUE)), file.size("b.fst"))
}
if (has_qs) {
  add_io("qs", "write", time_it(function() qs::qsave(dt_io, "b.qs")), file.size("b.qs"))
  add_io("qs", "read",  time_it(function() qs::qread("b.qs")), file.size("b.qs"))
}

con_io <- dbdt_connect("b.duckdb", quiet = TRUE)
add_io("dbdt (.duckdb)", "write", time_it(function() {
  as.dbdt(dt_io, conn = con_io, name = "bench", copy = TRUE, overwrite = TRUE)
}), file.size("b.duckdb"))
add_io("dbdt (.duckdb)", "read", time_it(function() {
  as.data.table(dbdt(con_io, "bench"))
}), file.size("b.duckdb"))
dbdt_disconnect(con_io)

io_df <- rbindlist(io_rows)
knitr::kable(io_df, col.names = c("Format", "Op", "Seconds", "Rows/sec", "File (MB)"))
```

```{r}
#| label: io-record
#| include: false
w <- io_df[format == "fwrite/fread" & op == "write", seconds]
r <- io_df[format == "fwrite/fread" & op == "read", seconds]
dw <- io_df[format == "dbdt (.duckdb)" & op == "write", seconds]
dr <- io_df[format == "dbdt (.duckdb)" & op == "read", seconds]
record("Round trip (2M rows)", "write vs fwrite", "fwrite()", w, dw)
record("Round trip (2M rows)", "read vs fread", "fread()", r, dr)
```

```{r}
#| label: io-cleanup
#| include: false
for (f in c("b.rds", "b.csv", "b.parquet", "b.fst", "b.qs", "b.duckdb")) {
  if (file.exists(f)) file.remove(f)
}
rm(dt_io); invisible(gc(FALSE))
```

## The honest cost: small, already-in-R data {#small-n}

[A file bigger than you want to load](#file-scan) and [Round-tripping a table to disk](#io) both move data of a size where paying DuckDB's dispatch and SQL-translation cost is worth it. It shouldn't be free at the other end -- a handful of rows already sitting in an R `data.table`, filtered through `dbdt`, has to beat plain `data.table[...]` doing the exact same thing with zero indirection. Here's that checked directly, at `n = 2,000`:

```{r}
#| label: bench-small-n
small <- make_data(2000)
con_small <- dbdt_example()
d_small <- as.dbdt(small, conn = con_small)
print(microbenchmark(
  data.table = small[category == "cat_05"],
  dbdt       = d_small[category == "cat_05"],
  times = 30
))
dbdt_disconnect(con_small)
```

Plain `data.table` wins here, and by design: v1 of `dbdt` is deliberately eager (see [Design notes](../README.md#design-notes)), so every `[` on a handle is a full round trip -- build SQL, send it to DuckDB, get rows back -- against `data.table`'s direct vector scan with no serialization at all. For data that already lives in R and comfortably fits, that's a real cost with no offsetting benefit. `dbdt` earns its keep once the data is bigger than you want to hold in R, already lives in a database, or the "already fits" assumption stops being true -- which is the whole point of [the section above](#file-scan).

## Summary

```{r}
#| label: tbl-summary
#| tbl-cap: !expr 'sprintf("Observed timings (this machine -- yours will vary): %d cores, %s.", parallel::detectCores(), R.version.string)'
summary_df <- do.call(rbind, results)
summary_df$baseline_s <- round(summary_df$baseline_s, 3)
summary_df$dbdt_s     <- round(summary_df$dbdt_s, 3)
summary_df$speedup    <- paste0(round(summary_df$speedup, 2), "x")
knitr::kable(
  summary_df[, -1],
  row.names = FALSE,
  col.names = c("Operation", "Baseline", "Baseline (s)", "dbdt (s)", "Speedup")
)
```

Read this table with the trade-off in view, not just the multiplier:

- **The pruning win is format-specific, and only shows up where it's real.** `dbdt_parquet()` beats a column-selecting `arrow::read_parquet()` because Parquet's layout gives DuckDB something to skip. `dbdt_csv()` loses to `fread()` on the exact same query against the exact same data, because CSV gives it nothing to skip -- there's no pruning to push down into a format that has no row groups or column stats. Same package, same operation, opposite result, because the file format is what actually decides it.
- **The round trip to disk is competitive, not dominant.** `dbdt`'s `.duckdb` file is in the same range as `fwrite`/`fst`/`arrow` for this mixed-string table -- worth it because you get a queryable database back, not a plain file, not because it's the fastest serializer on the shelf.
- **Small, already-in-memory data has no reason to go through SQL.** The overhead is real, it's consistent with the package's own design notes, and the fix isn't a faster `dbdt` -- it's using plain `data.table` for that case, which is what [the section above](#small-n) recommends.
