---
title: "Data checks & helpers"
description: "Profile a table, find duplicates, outliers and candidate keys, and small helpers for columns and sharing results."
---

```{r}
#| include: false
source("_setup.R")
```

Quick helpers for the first look at a table. Each takes a `data.table` (they stop with a clear message for anything else) and returns a `data.table` or prints a short summary, so the results compose with everything else you do.

The examples use `iris` and `airquality`, which has missing values:

```{r}
DT <- as.data.table(iris)
AQ <- as.data.table(airquality)
```

## Profile a table {#profile}

### `schema_dt()`: one row per column

Class, number of distinct non-missing values, and number of missing values. A quick overview of a wide table.

```{r}
schema_dt(AQ)
```

### `na_dt()`: missing values

The count and proportion of `NA`s in each column.

```{r}
na_dt(AQ)
```

### `freq_dt()`: the most common values

The top `n` values of one column (default 20), with counts and proportions, most frequent first. The column can be bare or a string; `n = Inf` returns every value.

```{r}
freq_dt(DT, Species)
```

```{r}
freq_dt(AQ, Month, n = 3)
```

### `spark_dt()`: a distribution at a glance

Buckets a numeric column into equal-width bins and prints a sparkline with the minimum, median and maximum. It is a quick look, not a plot: it only prints text, and the sparkline is also returned invisibly as a string.

```{r}
spark_dt(DT, Sepal.Length)
```

```{r}
spark_dt(AQ, Ozone, bins = 15)
```

### `key_dt()`: find candidate keys {#key-dt}

Looks for the smallest sets of columns that identify every row uniquely, which is what you want to know before a join. It reports *minimal* candidates: a combination is skipped if a smaller one already works. `cols` limits which columns are tried and `max_size` (default 3) how many are combined. The search grows combinatorially, so raise `max_size` carefully on wide tables.

```{r}
orders <- data.table(
  customer = c("a", "a", "b", "b", "c"),
  day      = as.Date("2024-03-01") + c(0, 1, 0, 0, 2),
  item     = c("x", "y", "x", "z", "x"),
  qty      = c(1, 2, 1, 5, 3)
)
key_dt(orders)
```

Candidates come from the data alone, so a coincidence can qualify, as `customer, qty` does here. Which one is meaningful is your call.

## Find problems {#problems}

### `dupe_dt()`: duplicate rows

Returns every row that belongs to a duplicate cluster (two or more rows sharing values across `by`, all columns by default), ordered by cluster and printed with group separators the way [`cdt()`](display.qmd#cdt) does. A `.dupe_group` column numbers the clusters. If nothing is duplicated you get a zero-row table.

```{r}
d <- data.table(
  id   = c(1, 1, 2, 3, 3, 3),
  name = c("a", "a", "b", "c", "d", "c")
)
dupe_dt(d)
```

Across all columns, row 5 (`3, d`) is not a duplicate. Matching on `id` alone makes it one:

```{r}
dupe_dt(d, by = "id")
```

### `outlier_dt()`: unusual values

Returns the rows where any of `cols` (all numeric columns by default) is an outlier, plus an `outlier_cols` column naming which column or columns triggered it.

- `method = "iqr"` (the default) flags values more than `threshold` times the interquartile range beyond the quartiles. `threshold` defaults to 1.5.
- `method = "zscore"` flags values whose absolute z-score exceeds `threshold`, which defaults to 3.

```{r}
outlier_dt(DT)
```

```{r}
outlier_dt(DT, cols = "Sepal.Width", method = "zscore", threshold = 2.5)
```

## Columns {#columns}

### `rename_dt()`: rename by reference

`renames` is a named character vector of `new_name = "old_name"`. The table is changed in place.

```{r}
DT2 <- copy(DT)
rename_dt(DT2, c(sepal_len = "Sepal.Length", sepal_wid = "Sepal.Width"))
names(DT2)
```

### `set_null()`: drop columns by reference

A shorthand for `set(x, j = cols, value = NULL)`.

```{r}
set_null(DT2, c("Petal.Length", "Petal.Width"))
names(DT2)
```

### `e()`: choose columns with an expression {#e}

Inside `j`, `e()` evaluates an expression that produces column names (or positions, or a logical vector) and returns those columns. It is handy for selecting columns by pattern.

```{r}
DT[, e(grep("Sepal", names(DT), value = TRUE))][1:3]
```

```{r}
DT[, e(grep("Sepal", names(DT), value = TRUE, invert = TRUE))][1:3]
```

```{r}
DT[, e(1:2)][1:3]
```

::: {.callout-warning}
## `e()` ignores the row filter in `i`
`e()` looks the table up by name and rebuilds the columns from the whole table, so a filter in `i` is not applied. To select columns *and* rows, use `.SD` with `.SDcols`.

```{r}
dt <- data.table(x = 1:3, y = 4:6)
dt[x == 1, e(c("x", "y"))]
```

```{r}
dt[x == 1, .SD, .SDcols = c("x", "y")]
```
:::

## Share results {#share}

### `md_dt()`: a markdown table

Prints a GitHub-flavoured pipe table for pasting into docs, pull requests and issues. `n` limits the rows and the text is also returned invisibly.

```{r}
md_dt(DT[1:4])
```

Which renders as:

```{r}
#| results: asis
md_dt(DT[1:4])
```

### `copy_dt()`: straight to the clipboard

Renders the same markdown table and copies it. It works out of the box on Windows and macOS; on Linux it needs `xclip` or `xsel`.

```r
copy_dt(DT[1:4])
```

`clip_fun` replaces the clipboard writer with any function that takes the text, which is also how to try it without touching your clipboard:

```{r}
copy_dt(DT[1:4], clip_fun = function(txt) cat("would copy", nchar(txt), "characters\n"))
```
