---
title: "Printing & sampling"
description: "Thousands separators, coloured class rows, group-aware sampling, row highlighting, and friendlier str() and dput() output."
---

```{r}
#| include: false
source("_setup.R")
```

Raw `data.table` output is fast and practical. When you are demoing results, reviewing a sampled cohort, debugging a mixed-type table or pasting console output into a document, small readability improvements add up. This page covers what changes how a table *looks*: the print mask, grouped sampling and display, row highlighting, and the `str()` and `dput()` masks. For checks that *find* things, see [Data checks & helpers](checks.qmd).

Every example uses a 15-row slice of `iris`, five rows per species, so the output stays short:

```{r}
DT <- as.data.table(iris)[c(1:5, 51:55, 101:105)]
```

## The print mask {#print-mask}

`enable_dt_print_thousands()` is switched on when you attach the package. It wraps `data.table`'s own printer and post-processes what it prints:

- thousands separators in the row numbers, with alignment kept stable
- a coloured class row (`<int>`, `<char>`, `<Date>`, ...)
- an `ncol:` header above the table
- separators and value colours for grouped output (see [below](#grouped-sampling))
- duplicated column names marked in red

```{r}
big <- data.table(row_id = 1:10000, value = round(rnorm(10000), 2))
big
```

Long tables print compactly: with the mask on, a table of more than 11 rows shows its first and last five rows with a `---` line between. Plain `data.table` prints up to 100 rows before it truncates. To see more, raise `topn`, either for one call, `print(DT, topn = 50)`, or for the session, `options(datatable.print.topn = 50)`. Grouped output (below) is the exception: it prints every row.

Duplicated column names are easy to miss; the mask marks them:

```{r}
data.table(x = 1, x = 1, y = 2)
```

### Tuning it

Call `enable_dt_print_thousands()` again with different arguments. Each call replaces the previous settings, and anything you leave out goes back to its default.

| Argument | Default | Effect |
|:---|:---|:---|
| `big.mark` | `","` | Thousands separator for row numbers, and for the row count in `str()` |
| `color` | `TRUE` | Colour the class row |
| `class_colors` | `NULL` | Named vector mapping class tokens (`"<num>"`) to `cli` colour helpers (`"col_cyan"`) |
| `show_ncol` | `TRUE` | Print `ncol: <n>` above the table |
| `color_group_values` | `TRUE` | In grouped output, colour the values within each group |
| `group_value_mode` | `"distinct"` | How values are coloured: `"distinct"` or `"similarity"` (see [Colour modes](#colour-modes)) |
| `similarity_max_distance`, `similarity_max_relative` | `2`, `0.30` | Thresholds for `"similarity"` mode |
| `force_color` | `TRUE` | Emit colour even where `cli` would not, by setting `cli.num_colors = 256` while printing |
| `group_sep_fmt` | `"--------- Group: %s"` | `sprintf()` format for the separator line above each group |

For example, your own colours for the class row, and no `ncol:` line:

```{r}
enable_dt_print_thousands(
  class_colors = c("<num>" = "col_cyan", "<char>" = "col_yellow", "<fctr>" = "col_magenta"),
  show_ncol = FALSE
)
DT[1:3]
```

```{r}
#| include: false
enable_dt_print_thousands()
```

`disable_dt_print_thousands()` removes the mask and goes back to plain `data.table` printing.

## Grouped sampling {#grouped-sampling}

`sample_dt()` has two modes.

**Without `group`** it samples `n` rows:

```{r}
sample_dt(DT, n = 4)
```

**With `group`** it samples `n` *groups* and returns every row of each chosen group, sorted by group. That keeps whole cohorts together, which is what you want when reviewing sampled records rather than sampled rows. `group` can be a bare column name or a string.

```{r}
sample_dt(DT, n = 2, group = Species)
```

Each group gets a separator line, and within a group repeated values share a colour, so a value that differs from its neighbours stands out.

### Ordering by commonality

With `sort_coverage = TRUE` (the default), rows within each group are ordered by how common their values are: rows sharing the most values with the rest of the group come first. Here the `Kyle Haynes` rows are the group's common pattern; the others are variations.

```{r}
cohort <- data.table(
  cohort_id  = c(1, 1, 1, 1, 1, 2, 2, 2),
  given_name = c("Kyle", "Cyle", "Kyle", "Kylie", "Kyle", "Sarah", "Sara", "Sarah"),
  last_name  = c("Haynes", "Hines", "Haynes", "Hines", "Haynes", "Smith", "Smyth", "Smith")
)

sample_dt(cohort, n = 2, group = cohort_id, sort_coverage = FALSE)
```

```{r}
sample_dt(cohort, n = 2, group = cohort_id, sort_coverage = TRUE)
```

### Other arguments

| Argument | Default | Effect |
|:---|:---|:---|
| `color` | session default (`TRUE`) | Colour the output. `FALSE` gives plain output. |
| `sort_coverage` | `TRUE` | Order rows by commonality within each group (needs `group`) |
| `color_threshold` | `500L` | If any group has more rows than this, colour is switched off for the result |

## Colouring a table without sampling: `cdt()` {#cdt}

`cdt()` applies the same display treatment to a table you already have, with no sampling. Without `group` the whole table is one display unit and prints compactly, like any long table (the five middle rows are left out here). With `group` you get the separators and per-group colours, and every row is printed.

```{r}
cdt(DT)
```

```{r}
cdt(DT, group = Species)
```

`cdt()` takes the same `color`, `sort_coverage` and `color_threshold` arguments as `sample_dt()`. Because grouped output prints every row, narrow a large table before you group it.

## Highlighting rows {#highlight}

`highlight_dt()` colours entire rows that match a condition, so anomalies show up in ordinary console output. The condition is evaluated inside the table (or is a logical vector as long as the table). `color` names a `cli` colour helper, `"col_red"` by default; `"red"` works too.

```{r}
highlight_dt(DT, Sepal.Length > 6.5)
```

```{r}
highlight_dt(DT, Species == "setosa", color = "col_cyan")
```

::: {.callout-note}
`data.table` prints only the first and last few rows of a long table, and highlighting can only colour rows that are printed. To make sure the rows you care about are on screen, filter first or raise `options(datatable.print.topn)`.
:::

## Colour modes {#colour-modes}

In grouped output, values are coloured within each group and column. There are two ways to decide which values share a colour:

- **distinct** (the default): every distinct value gets its own colour, so equal values match and any difference shows.
- **similarity**: strings that are close to each other share a colour, which makes typos and near-duplicates easy to see. Two values are close when their edit distance is at most `similarity_max_distance`, or at most `similarity_max_relative` of the longer value's length, ignoring case and punctuation.

A mode stored on a table wins over the session setting, and `sample_dt()`, `cdt()` and `dupe_dt()` store `"distinct"` on their results. So `group_value_mode = "similarity"` does not change how the output of those three prints; see [Limitations](limitations.qmd#colour-mode).

## Switching colour on and off {#colour-default}

`switch_col()` sets the session default for the `color` argument of `sample_dt()`, `cdt()` and `dupe_dt()`:

```{r}
switch_col(FALSE)
sample_dt(DT, n = 3)
```

```{r}
switch_col(TRUE)
```

Pass `color = TRUE` or `color = FALSE` to a single call to override the default just once.

## `str()` and `dput()` {#str-dput}

`enable_dt_str_mask()` gives `str()` a data.table-shaped header, shows every column, and drops the `.internal.selfref` line. Other objects go to the usual `str()` untouched.

```{r}
str(DT)
```

`enable_dt_dput_mask()` removes the `.internal.selfref` pointer from `dput()` output, so what you paste into an issue or a script runs as it is:

```{r}
dput(DT[1:2])
```

Without the masks:

```{r}
disable_dt_str_mask()
disable_dt_dput_mask()
str(DT)
dput(DT[1:2])
```

```{r}
turn_everyone_on()
```

## All the switches {#switches}

| Function | What it does |
|:---|:---|
| `turn_everyone_on()` | Turns on the print, `str()` and `dput()` masks, and colour for `sample_dt()`, `cdt()` and `dupe_dt()` |
| `enable_dt_print_thousands()`, `disable_dt_print_thousands()` | The print mask |
| `enable_dt_str_mask()`, `disable_dt_str_mask()` | The `str()` mask |
| `enable_dt_dput_mask()`, `disable_dt_dput_mask()` | The `dput()` mask |
| `switch_col()` | The session default for colour in `sample_dt()`, `cdt()` and `dupe_dt()` |

The masks live in your global environment; see [Get started](getting-started.qmd#attaching-the-package).
