---
title: "Query with [i, j, by]"
description: "How data.table's i, j and by become one SQL query, what is translated, and what is not."
---

```{r}
#| include: false
source("_setup.R")
```

On a `duckdt` handle, `d[i, j, by]` translates the three slots into one SQL `SELECT`, runs it inside DuckDB, and returns a `data.table`. The slots mean what they always do: `i` says which rows, `j` says what to return, `by` says grouped how.

```{r}
con <- dbdt_example()
orders <- dbdt(con, "orders")
items <- dbdt(con, "items")
customers <- dbdt(con, "customers")
```

## The three slots {#slots}

| You write | DuckDB runs (simplified) |
|:---|:---|
| `items[price > 100]` | `SELECT * FROM items WHERE price > 100` |
| `items[price > 100, .(label, price)]` | `SELECT label, price FROM items WHERE price > 100` |
| `orders[, .(n = .N), by = customer_id]` | `SELECT customer_id, count(*) AS n FROM orders GROUP BY customer_id` |

```{r}
items[price > 100]
```

```{r}
items[price > 100, .(label, price)]
```

```{r}
orders[, .(n = .N), by = customer_id][order(customer_id)]
```

Every `[` runs **immediately**: one query, one `data.table` back. That keeps the mental model identical to plain `data.table`, and you keep chaining on the result the way you always would, as the `[order(customer_id)]` above does. The result has left the database at that point; to keep a subset *in* the database, use [`dbdt_temp()`](duckdt-join-write.qmd#temp).

## Filtering rows: `i` {#i}

| Kind | Supported |
|:---|:---|
| Comparison | `==`, `!=`, `<`, `<=`, `>`, `>=` |
| Logic | `&`, `\|`, `!` (and `&&`, `\|\|`), parentheses |
| Membership and range | `%in%`, `%chin%`, `%notin%`, `%between%`, `between()` |
| Text | `%like%`, `%ilike%`, `%flike%`, `%plike%`, `grepl()`, `startsWith()`, `endsWith()` |
| Missing values | `is.na()` |
| Arithmetic | `+`, `-`, `*`, `/`, `^`, `%%`, `%/%` |
| Functions | the ones under [Functions](#functions) |

```{r}
items[price %between% c(5, 200)]
```

```{r}
orders[customer_id %in% c(1, 3) & !is.na(ordered_on)]
```

### Text matching {#text}

Like `data.table`, the `%like%` family matches **regular expressions**, not SQL `LIKE` wildcards, except `%flike%`, which is a literal substring match. They run as DuckDB's `regexp_matches()` and `contains()`.

| Operator | Matches |
|:---|:---|
| `%like%` | regular expression, case-sensitive |
| `%ilike%` | regular expression, case-insensitive |
| `%flike%` | literal substring |
| `%plike%` | alias of `%like%` |

```{r}
customers[name %ilike% "^a"]
```

```{r}
customers[name %flike% "Turing"]
```

DuckDB's regular expressions use RE2, which has no PCRE backreferences or lookaround, so `%plike%` is a best-effort alias for `%like%` and not real Perl-regex support.

### Using R values {#r-values}

A name that is not a column is looked up in the calling environment, as it is in `data.table`, so variables and function arguments work:

```{r}
min_price <- 100
items[price > min_price, .(label, price)]
```

```{r}
pricier_than <- function(threshold) items[price > threshold, .(label)]
pricier_than(50)
```

Only *values* travel this way. A call to an R function is not evaluated in R: the functions under [Functions](#functions) are translated, and anything else is sent to DuckDB **by name**, so it fails if DuckDB has no such function. Work out R-side values first, then refer to them by name:

```{r}
#| error: true
cutoff_for <- function(month) as.Date(paste0("2024-", month, "-05"))
orders[ordered_on >= cutoff_for("03")]
```

```{r}
cutoff <- cutoff_for("03")
orders[ordered_on >= cutoff]
```

## Choosing and computing columns: `j` {#j}

`j` is a bare column name, or `.(...)` / `list(...)` to select, rename and compute. It always returns a `data.table`, even for a single column.

```{r}
items[, .(label, price_incl_gst = price * 1.1)]
```

Aggregates work inside `.()`, and `.N` is the row count:

```{r}
orders[, .(n = .N, first_order = min(ordered_on), last_order = max(ordered_on))]
```

A bare call in `j` is not supported. Wrap it in `.()`:

```{r}
#| error: true
items[, mean(price)]
```

```{r}
items[, .(avg_price = mean(price))]
```

`j` can also be a `:=` assignment, which writes to the table; see [Join, write & merge](duckdt-join-write.qmd#write).

### Functions {#functions}

These R and `data.table` functions are translated, so they mean in a query what they mean in R. A `pkg::` prefix is fine (`data.table::fifelse()`).

| Kind | Functions |
|:---|:---|
| Aggregates | `sum()`, `mean()`, `median()`, `min()`, `max()`, `sd()`, `var()`, `length()`, `uniqueN()`, `.N` |
| Conditions | `ifelse()`, `fifelse()` (with its `na` argument), `fcase()` (with `default`), `fcoalesce()`, `between()` |
| Text | `paste0()`, `paste()` (with `sep`), `substr()`, `substring()`, `nchar()`, `toupper()`, `tolower()`, `trimws()`, `startsWith()`, `endsWith()`, `grepl()` (with `ignore.case` and `fixed`) |
| Numbers | `abs()`, `round()`, `trunc()`, `log()` (with `base`), `log2()`, `log10()`, `pmin()`, `pmax()` |
| Types | `as.numeric()`, `as.double()`, `as.integer()`, `as.character()`, `as.Date()`, `as.IDate()` |
| Dates | `year()`, `month()`, `quarter()`, `mday()`, `wday()`, `yday()`, `hour()`, `minute()`, `second()`, as in `data.table` (`wday()` counts from Sunday = 1) |

Any other call is sent to DuckDB by name, upper-cased, so any DuckDB function you can write in R call syntax works too: `coalesce()`, `regexp_replace()`, `date_trunc()` and so on.

```{r}
customers[, .(initial = substr(name, 1, 1), shout = toupper(name))]
```

```{r}
items[, .(label, band = fcase(price < 10, "cheap", price < 100, "mid", default = "dear"))]
```

```{r}
orders[, .(n = .N, customers = uniqueN(customer_id)), by = .(year = year(ordered_on), q = quarter(ordered_on))]
```

A comparison can be summed or averaged, as in R: `mean(price > 100)` is the share of rows over 100.

A few results differ from R in ways that follow from SQL:

- `uniqueN()` doesn't count a missing value as a value, so it matches `uniqueN(x, na.rm = TRUE)`.
- `paste0()` and `paste()` turn a missing value into `""`, where R writes `"NA"`, and DuckDB writes a whole double such as `3` as `3.0`.
- `paste(collapse = )` combines rows rather than columns and isn't translated.

## Grouping: `by` {#by}

`by` takes a bare column, a character vector, or `.()`, `list()` or `c()`. Give a computed group a name and it becomes a column of the result:

```{r}
orders[, .(n = .N), by = .(year = year(ordered_on), customer_id)][order(customer_id)]
```

```{r}
items[, .(n = .N, avg_price = mean(price)), by = .(pricey = price > 100)][order(pricey)]
```

## Missing values {#na}

`%in%` behaves as in R and never returns `NA`. SQL aggregates already skip `NULL`, so `mean(x)` is the same as R's `mean(x, na.rm = TRUE)`, whether or not you write `na.rm = TRUE`. That argument is accepted and dropped:

```{r}
customers[, .(mean_id = mean(customer_id, na.rm = TRUE))]
```

`ifelse()` and `fifelse()` keep R's rule that a missing test gives a missing result, rather than SQL's habit of falling through to the "no" branch.

## Ordering and row positions {#ordering}

DuckDB tables have no inherent row order. `order()` in `i` sorts in the database, as `ORDER BY`: `-x` or `decreasing = TRUE` sort descending, and missing values go last, as in R.

```{r}
items[order(-price), .(label, price)]
```

Sorting a query in `i` is all-or-nothing, since `i` can hold either a filter or a sort, not both. To filter *and* sort, filter in the database and sort the result, which is an ordinary `data.table`:

```{r}
items[price > 5, .(label, price)][order(-price)]
```

`order()` can't be combined with `by` (sort the grouped result instead), and row positions such as `d[1:5]` aren't supported; use `head()`. Without an `order()`, expect no particular row order from any query.

## Peeking and sampling {#peeking}

```{r}
head(items, 2)
```

```{r}
tail(items, 2)
```

```{r}
dbdt_sample(items, 2)
```

| Function | Runs as |
|:---|:---|
| `head(d, n)` | `SELECT * ... LIMIT n` |
| `tail(d, n)` | `... LIMIT n OFFSET (rows - n)` |
| `dbdt_sample(d, n)` | `... USING SAMPLE reservoir(n ROWS)` |

A negative `n` works as in base R: `head(d, -2)` is everything but the last two rows. `tail()` reflects DuckDB's current scan order and not a guaranteed original order. `dbdt_sample()` samples without replacement; DuckDB's sampling clause has no replacement option. It is a plain function because base `sample()` is not an S3 generic, so `sample(d, n)` would never reach a method.

`nrow()`, `ncol()`, `dim()`, `names()` and `colnames()` also work on a handle.

## Profiling a table {#profile}

`dbdt_profile()` is the database-side [`schema_dt()`](checks.qmd#profile): for every column, the number and share of missing values, the number of distinct values, and the smallest and largest. The database computes all of it in one query, and only the summary comes back, which makes it the way to get to know a table too big to pull into R.

```{r}
dbdt_profile(orders)
```

Counting distinct values is the expensive part on a large table; `distinct = FALSE` skips it. `cols =` profiles just some columns.

## Seeing the SQL {#sql}

`dbdt_sql()` takes the same `i`, `j` and `by` as `[` and returns the SQL instead of running it, to check a translation or to take a slow query somewhere that shows its plan (DuckDB's `EXPLAIN ANALYZE`, or SQL Server Management Studio):

```{r}
dbdt_sql(orders, !is.na(ordered_on), .(n = .N, last = max(ordered_on)), by = customer_id)
```

To watch every statement `duckdt` sends, catalogue lookups included, with the time each took, turn on tracing:

```{r}
options(duckdt.trace = TRUE)
orders[customer_id == 1, .(n = .N)]
options(duckdt.trace = NULL)
```

## Binary columns {#binary}

No driver hands a binary column (`BLOB` and `BIT` on DuckDB; `varbinary`, `binary` and `image` on SQL Server) back as an R vector, and asking for one fails the *whole* query, so a single such column would make a table unreadable. When rows are on their way into R, `duckdt` leaves binary columns out, and says so.

```{r}
#| output: false
DBI::dbExecute(con, "CREATE TABLE files (id INTEGER, name VARCHAR, payload BLOB)")
DBI::dbExecute(con, "INSERT INTO files VALUES (1, 'a.bin', 'abc'::BLOB)")
files <- dbdt(con, "files")
```

```{r}
files
```

```{r}
head(files)
```

Name the column in `j` to fetch it anyway (it fails if the driver cannot convert it). Anything that stays in the database, such as `:=`, [`dbdt_temp()`](duckdt-join-write.qmd#temp) and [`dbdt_merge()`](duckdt-join-write.qmd#merge), still sees every column. You can also reduce a binary column inside the database, here to its size in bytes:

```{r}
files[, .(name, bytes = octet_length(payload))]
```

## What is not supported {#not-supported}

- **Lazy, chained queries.** Every `[` runs at once. [`dbdt_temp()`](duckdt-join-write.qmd#temp) is the escape hatch: it runs a query into a temporary table and hands back a handle.
- **Row positions in `i`, and `order()` with `by`**, as above.
- **Grouped `:=`**, that is `by =` together with a write.
- **`dbdt_sample(replace = TRUE)`.**
- **R functions without a translation or a DuckDB namesake**, such as `format()` or your own functions. Compute R-side values first ([above](#r-values)).

See [Limitations](limitations.qmd) for the full list, including the caveats that apply to writes and joins.

```{r}
#| include: false
dbdt_disconnect(con)
```
