---
title: "Connect & load data"
description: "Open a DuckDB database, and bring in data from R, from files, or from tables that already exist."
---

```{r}
#| include: false
source("_setup.R")
# The file-based examples below write to the working directory; keep that in a scratch folder.
knitr::opts_knit$set(root.dir = tempdir())
```

`dbdt()` lets you write `data.table`'s `d[i, j, by]` against a database. The steps are always the same: connect, get a *handle* on some data, query it. This page covers the first two; [Query](duckdt-query.qmd) covers the third.

## How it fits together

```{mermaid}
%%| echo: false
%%| fig-width: 9
flowchart LR
  A["Your R code<br/>d[state == 'WA', .N, by = locality]"] --> B["duckdt<br/>translates to SQL"]
  B --> C["DuckDB<br/>does the work"]
  C --> D["A data.table<br/>just the answer"]
  E[(".duckdb file<br/>.parquet / .csv")] --- C
```

A `duckdt` handle is a pointer: *this connection, that table*. No data is copied and none is loaded into R. Every `[` sends one SQL query to DuckDB and returns the result as an ordinary `data.table`, so the rest of your R code does not change.

## Connect {#connect}

`dbdt_connect()` opens a DuckDB database. With no arguments it is an in-memory scratch database that disappears when you disconnect.

```{r}
con <- dbdt_connect()
```

The summary says what you just opened and gives the two calls that get you moving. Arguments:

| Argument | Default | Effect |
|:---|:---|:---|
| `dbdir` | `":memory:"` | Path to a database file. A file that does not exist yet is created. Forward slashes (`"C:/data/warehouse.duckdb"`) work on Windows. |
| `read_only` | `FALSE` | Open the file read-only. Good for a database you only mean to look at, and required if another process has it open. |
| `quiet` | `FALSE` | Skip the summary. |
| `...` | | Passed to `duckdb::duckdb()`. |

`dbdt_connect()` is a thin wrapper over `DBI::dbConnect(duckdb::duckdb(dbdir = ...))`. Every function in the package takes a plain DBI connection, so opening one yourself is equally fine.

Close it when you are done. This also shuts the database down, which flushes a file-backed database cleanly:

```{r}
dbdt_disconnect(con)
```

## An example database {#example}

`dbdt_example()` builds a small in-memory database of a toy shop, with real primary and foreign keys declared. The docs use it throughout.

```{r}
con <- dbdt_example()
dbdt_tables(con)
```

`customers`, `items`, `orders` and `order_lines`; `order_lines` has a compound primary key and points at both `orders` and `items`. `dbdt_tables()` and its relatives are covered in [Explore a database](duckdt-explore.qmd).

## Getting data in {#getting-data-in}

```{mermaid}
%%| echo: false
%%| fig-width: 9
flowchart TD
  A["What have you got?"] --> B["A data.frame<br/>in R"]
  A --> C["A .csv or .parquet<br/>file"]
  A --> D["A table already<br/>in the database"]
  B --> B1["as.dbdt(x)<br/>zero-copy view"]
  B --> B2["as.dbdt(x, copy = TRUE)<br/>written into DuckDB"]
  C --> C1["dbdt_csv(path)<br/>dbdt_parquet(path)"]
  D --> D1["dbdt(con, 'name')"]
```

### From R: `as.dbdt()` {#as-duckdt}

By default this is **zero-copy**: the R object is registered as a virtual view inside DuckDB, with nothing serialised. The handle is read from R's memory when you query it.

```{r}
d <- as.dbdt(mtcars, conn = con)
d
```

Pass `copy = TRUE` to physically write the data into a DuckDB table instead. You need this for a table you want to change with `:=` or [`dbdt_merge()`](duckdt-join-write.qmd), and for anything that must survive the session.

```{r}
d <- as.dbdt(mtcars, conn = con, name = "mtcars_tbl", copy = TRUE)
d
```

Other arguments: `name` (the table or view name in DuckDB, defaulting to the object's name), `overwrite = TRUE` to replace one that exists, and `conn`. With no `conn` a new in-memory database is created for you.

::: {.callout-note}
## Row names are not carried over
`mtcars` keeps its car names in the row names, and DuckDB has no row names, so they are dropped. Put them in a column first, so they exist in the table:

```{r}
cars <- data.table(car = rownames(mtcars), mtcars)
d <- as.dbdt(cars, conn = con, copy = TRUE)
names(d)[1:4]
```
:::

#### List-columns

Against DuckDB, R list-columns round-trip through DuckDB's native `LIST` type, both zero-copy and with `copy = TRUE`, and come back as a list-column with no extra steps:

```{r}
x <- data.table(id = 1:2, tags = list(c("a", "b"), "c"))
dl <- as.dbdt(x, conn = con, name = "with_tags", copy = TRUE)
as.data.table(dl)$tags
```

As with DuckDB's `LIST` type, every element of one list-column must coerce to a single atomic type, so a column that mixes an integer-vector cell with a character-vector cell cannot be represented. List-columns are not supported on SQL Server.

### From files: `dbdt_csv()` and `dbdt_parquet()` {#files}

These create a lazy view over one or more files. The file is not read into R, and DuckDB reads only what a query needs, out of core, so this works on files bigger than memory.

```{r}
fwrite(iris, "iris.csv")
v <- dbdt_csv("iris.csv", conn = con)
v
```

`path` can be several files, and Parquet paths can use wildcards, such as `dbdt_parquet("data/*.parquet")`. The view is named after the file unless you give `name`; here that would be `customers`, which is already a table in this database (DuckDB will not replace a table with a view), so the name is given explicitly.

```{r}
#| output: false
DBI::dbExecute(con, "COPY customers TO 'customers.parquet' (FORMAT PARQUET)")
```

```{r}
p <- dbdt_parquet("customers.parquet", conn = con, name = "customers_pq")
p
```

Both are DuckDB-only (they use DuckDB's `read_csv_auto` and `read_parquet`), and a file-backed view is read-only.

### A table that already exists: `dbdt()` {#duckdt}

`dbdt(conn, table)` wraps a table or view that is already in the database. Nothing is copied.

```{r}
d <- dbdt(con, "customers")
d
```

`nrow()`, `ncol()`, `dim()` and `names()` work on a handle and run tiny queries. Handles made this way are **read-only by default**, which the print flags: `:=` and `dbdt_merge()` refuse to run against them until you opt in with `dbdt(con, "customers", writable = TRUE)`. The point is to stop you changing a table you only meant to look at, for example after reconnecting to a file.

### Reading a handle {#status}

What a handle's first line tells you:

| Print shows | What it is | `:=` and `dbdt_merge()` |
|:---|:---|:---|
| `<duckdt> t [3 x 3]` | A real table, opened writable (`as.dbdt(copy = TRUE)` or `writable = TRUE`) | Allowed |
| `... (read-only)` | A real table wrapped with `dbdt()` | Refused until `writable = TRUE` |
| `... (view)` | A view: zero-copy `as.dbdt()`, `dbdt_csv()`, `dbdt_parquet()` | Never; views cannot be written |
| `... (temp)` | A temporary table from [`dbdt_temp()`](duckdt-join-write.qmd#temp) | Allowed, on the temporary copy |

## A database that outlives the session {#persist}

By default `duckdt` is in-memory. To keep data, connect to a file and write with `copy = TRUE`. This is the whole round trip, from a `data.table` in memory to a file, and back in a later session:

```{r}
sales <- data.table(
  order_id = 1:5,
  customer = c("Alice", "Bob", "Alice", "Carol", "Bob"),
  amount   = c(120.50, 89.00, 45.25, 300.00, 15.75)
)

# Session 1: write it to a file.
file_con <- dbdt_connect("sales.duckdb")
as.dbdt(sales, conn = file_con, name = "sales", copy = TRUE)
dbdt_disconnect(file_con)
```

```{r}
# Session 2 (or later): open the same file and pick up where you left off.
file_con <- dbdt_connect("sales.duckdb")
dbdt_tables(file_con)
```

```{r}
d <- dbdt(file_con, "sales")
d[, .(total = sum(amount), n = .N), by = customer][order(customer)]
```

`copy = TRUE` is what makes the table persistent; the zero-copy default is a view over your R object and disappears with the session. Use `overwrite = TRUE` if you re-run the write against a file that already has the table. The handle from `dbdt()` is read-only by default, as above; to write back, reopen it with `writable = TRUE` (see [Join, write & merge](duckdt-join-write.qmd#write)).

```{r}
dbdt_disconnect(file_con)
```

Always disconnect from a file-backed database: shutting it down flushes DuckDB's write-ahead log and closes the file cleanly. `dbdt_disconnect(con)` does this; so does `DBI::dbDisconnect(con, shutdown = TRUE)`.

## Getting data out {#getting-data-out}

Every `[` already returns a `data.table`. To pull an entire table into R:

```{r}
as.data.table(dbdt(con, "items"))
```

That is `SELECT *` materialised in R, so use it on tables that fit. For a large table, [query it](duckdt-query.qmd) instead, or peek with `head()`, `tail()` and `dbdt_sample()`.

Binary columns (`BLOB` and `BIT` on DuckDB) cannot be converted to R and are left out of these results, with a note; see [Query](duckdt-query.qmd#binary).

```{r}
#| include: false
dbdt_disconnect(con)
```
