---
title: "Join, write & merge"
description: "Join tables inside the database, keep a subset there with temporary tables, and change tables in place with := and dbdt_merge()."
---

```{r}
#| include: false
source("_setup.R")
```

Three verbs, and one rule of thumb: **`dbdt_join()` reads, `dbdt_merge()` writes, and `dbdt_temp()` keeps a subset in the database** so the other two can use it.

| I want to... | Use | Changes the inputs? |
|:---|:---|:---|
| Combine two tables into a new result | [`dbdt_join()`](#join), or `merge()` | No |
| Keep a filtered subset in the database | [`dbdt_temp()`](#temp) | No (it creates a temporary table) |
| Add or change a column | [`:=`](#write) | Yes, in place |
| Patch a table with corrections: update, insert, delete | [`dbdt_merge()`](#merge) | Yes, in place |

The examples use the toy shop database:

```{r}
con <- dbdt_example()
orders <- dbdt(con, "orders")
customers <- dbdt(con, "customers")
items <- dbdt(con, "items")
```

## Joining: `dbdt_join()` {#join}

`dbdt_join()` is a database-side `merge()`: the two tables are joined in SQL, and only the result comes back to R. Both inputs are left alone.

```{r}
dbdt_join(orders, customers, by = "customer_id")
```

Either side may be a `"duckdt"` handle or an ordinary `data.frame` or `data.table`. An R-side table is copied into a temporary table on the connection first, so the join itself always runs in the engine.

```{r}
regions <- data.table(city = c("Perth", "Sydney"), region = c("West", "East"))
dbdt_join(customers, regions, by = "city")
```

`merge(x, y, ...)` on a handle calls `dbdt_join()`, so the familiar spelling works too:

```{r}
merge(orders, customers, by = "customer_id")
```

The arguments behave as in base `merge()`:

| Argument | Effect |
|:---|:---|
| `by` | Join column(s), present in both. Defaults to every column the two share (a natural join). |
| `by.x`, `by.y` | Join on differently named columns; give both. The result keeps `by.x`'s names. |
| `all`, `all.x`, `all.y` | No `all*` is an inner join; `all.x` is a left join, `all.y` a right join, `all` a full outer join. |
| `suffixes` | Disambiguate non-key columns both sides have. Default `c(".x", ".y")`. |
| `sort` | Sort by the join columns (default `TRUE`). `FALSE` returns the engine's order, which is faster and arbitrary. |

Alan Turing has no orders. A left join from `customers` keeps him, with `NA` where the order columns would be:

```{r}
dbdt_join(customers, orders, by = "customer_id", all.x = TRUE)
```

Joining on differently named columns:

```{r}
discounts <- data.table(id = 1:3, discount = c(0, 0.1, 0.2))
dbdt_join(items, discounts, by.x = "item_id", by.y = "id")
```

At least one side must be a `"duckdt"` handle, since that says which database to run in, and if both are, they must share a connection. Joining two R data frames is `data.table`'s job. Missing keys never match in a database join; see [Limitations](limitations.qmd#na-keys).

## Keeping a subset in the database: `dbdt_temp()` {#temp}

`d[i, j, by]` returns a `data.table`, so the rows have left the database. `dbdt_temp()` runs the same query, leaves the result in the database as a temporary table, and hands back a `"duckdt"` handle. Nothing crosses into R.

```{r}
sub <- dbdt_temp(orders, customer_id == 1)
sub
```

The handle works anywhere a handle does. Here the join runs entirely inside DuckDB:

```{r}
dbdt_join(sub, customers, by = "customer_id")
```

Drop it when you are finished:

```{r}
dbdt_drop(sub)
```

- `i`, `j` and `by` are all optional; `dbdt_temp(x)` copies the whole table. `name` sets the table name, which is otherwise generated and unique within the session.
- The handle is materialised and writable, so `:=` and `dbdt_merge()` work on it. Writes land on the temporary copy and leave the source table alone.
- On DuckDB it is a genuine `TEMP` table in the session's `temp` schema, and it disappears when the connection closes. On SQL Server it is created with `SELECT ... INTO` as an ordinary table on the current schema, so it outlives the session and you should `dbdt_drop()` it.
- `dbdt_drop()` refuses handles that `dbdt_temp()` did not create, unless you pass `force = TRUE`.
- Use the handle `dbdt_temp()` returns. Wrapping the temporary table again with `dbdt()` loses its temporary status.

```{r}
#| error: true
dbdt_drop(orders)
```

## Writing: `:=` {#write}

`:=` adds or changes a column **in the database**. It needs a handle over a real table that is writable. `as.dbdt(x, copy = TRUE)` gives you one immediately, since you just created that table:

```{r}
cars <- as.dbdt(data.table(car = rownames(mtcars), mtcars), conn = con, name = "cars", copy = TRUE)
```

```{r}
#| output: false
cars[, kw := hp * 0.7457]        # add a computed column
cars[cyl == 6, kw := kw * 1.1]   # change matching rows only
cars[, `:=`(wt_kg = wt * 453.6, mpg_l100 = 235.215 / mpg)]   # several at once
```

```{r}
cars[cyl == 6, .(car, hp, kw, wt_kg)][order(car)][1:3]
```

`:=` returns the handle, so at the console it prints after each assignment; that output is hidden in the chunk above. Wrap a call in `invisible()` if you would rather it stayed quiet.

A handle from `dbdt(con, "table")` is **read-only by default**, so a table you wrapped only to look at cannot be changed by accident. Opt in when you mean to write:

```{r}
#| error: true
customers[, city_upper := toupper(city)]
```

```{r}
#| output: false
ol <- dbdt(con, "order_lines", writable = TRUE)
ol[, qty_doubled := qty * 2L]
```

```{r}
ol[, .(order_id, line_no, qty, qty_doubled)][order(order_id, line_no)]
```

A view (from a zero-copy `as.dbdt()`, or `dbdt_csv()` and `dbdt_parquet()`) can never be written to.

### How `:=` writes {#write-how}

`:=` changes the table **in place**. A column that exists is set with `UPDATE` (with `i` as its `WHERE` clause, so only the matching rows are touched); a new column is added with `ALTER TABLE ... ADD` and then filled in; `col := NULL` drops a column, as in `data.table`. The table keeps its primary key, its constraints and its indexes:

```{r}
#| output: false
DBI::dbExecute(con, "CREATE TABLE tagged (id INTEGER PRIMARY KEY, x INTEGER NOT NULL)")
DBI::dbExecute(con, "INSERT INTO tagged VALUES (1, 10), (2, 20)")

constraints <- function(tbl) {
  DBI::dbGetQuery(con, sprintf(
    "SELECT constraint_type FROM duckdb_constraints() WHERE table_name = '%s'", tbl
  ))$constraint_type
}
```

```{r}
tagged <- dbdt(con, "tagged", writable = TRUE)
invisible(tagged[, y := x * 2L])
invisible(tagged[id == 1, x := 11L])
constraints("tagged")
tagged[order(id)]
```

All the assignments in one `:=` call are translated before anything is written, and written in one transaction, so they land together or not at all. A later assignment in the call sees an earlier one's result, `` `:=`(kw = hp * 0.7457, kw2 = kw * 2) ``.

Types follow `data.table`. With `i`, the column keeps its type and the new values are converted to it; assigning a whole column replaces it, type and all, so `n := n / 2` on an integer column gives a double one. A new column from an R value gets R's type: `:= 0.5` is a double, `:= "no"` is text with room for longer values later.

Because nothing is dropped and recreated, `:=` also works on a table that other tables reference, such as `customers` here, which `orders` points at:

```{r}
cust <- dbdt(con, "customers", writable = TRUE)
invisible(cust[, city_upper := toupper(city)])
cust[, .(customer_id, city, city_upper)][order(customer_id)]
```

`:= NULL` drops a column, though DuckDB won't drop one from a table others reference, like `customers`:

```{r}
invisible(tagged[, y := NULL])
names(tagged)
```

There is no grouped `:=` (`by` together with a write). On SQL Server, see [what differs there](duckdt-mssql.qmd#dialect).

## Merging: `dbdt_merge()` {#merge}

`dbdt_merge()` patches a table **in place** with a subset. The subset `y` (a data frame, or another `"duckdt"` on the same connection) is staged into a temporary table first, so the big table never round-trips through R. Then, all in SQL:

- rows in both are **updated** (`update = TRUE`, the default),
- rows only in `y` are **inserted** (`insert = TRUE`, the default),
- rows only in `x` are **deleted** (`delete = TRUE`; off by default).

```{r}
kv <- as.dbdt(data.table(k = 1:3, v = c("a", "b", "c")), conn = con, name = "kv", copy = TRUE)
patch <- data.table(k = c(2L, 9L), v = c("B", "Z"))

dbdt_merge(kv, patch, by = "k")     # update k = 2, insert k = 9
kv[, .(k, v)][order(k)]
```

```{r}
# Update only: skip rows that are not already there.
dbdt_merge(kv, data.table(k = c(3L, 10L), v = c("C", "X")), by = "k", insert = FALSE)
kv[, .(k, v)][order(k)]
```

```{r}
# Replace the subset entirely: also delete rows that are not in `y`.
dbdt_merge(kv, data.table(k = c(1L, 2L), v = c("one", "two")), by = "k", delete = TRUE)
kv[, .(k, v)][order(k)]
```

- **`x`** must be a `"duckdt"` handle over a materialised, writable table, since the merge changes it. A `[` result has already left the database and cannot be used here.
- **`by`** defaults to every column `x` and `y` share, like base `merge()`. A column cannot be both a match key and something `update` sets from `y`, so pass `by =` explicitly whenever `y` carries value columns you want updated.
- Columns of `y` that are not in `x` are ignored. Add the column with `:=` first.
- On DuckDB the merge is an `UPDATE ... FROM`, an anti-join `INSERT` and, with `delete = TRUE`, an anti-join `DELETE`, wrapped in a transaction. On SQL Server it is a single native `MERGE` statement.

`dbdt_merge()` writes. To combine two tables into a new result without touching either, use [`dbdt_join()`](#join).

```{r}
#| include: false
dbdt_disconnect(con)
```
