---
title: "Get started"
description: "Install the package, see what attaching it changes, and take a first step in each half."
---

```{r}
#| include: false
source("_setup.R")
```

## Install

```r
# From GitHub
devtools::install_github("KyleHaynes/data.table.ext")

# Or from a local checkout
install.packages(".", repos = NULL, type = "source")
```

### Optional packages {#optional-packages}

`data.table`, `cli`, `DBI` and `duckdb` are required and installed with the package. Everything else is only needed for the feature named:

| Package | Needed for |
|:---|:---|
| `DiagrammeR` | `dbdt_dm_render()`, and printing the result of `dbdt_dm_dot()` |
| `DiagrammeRsvg`, `rsvg` | `dbdt_dm_export()` to image files (`rsvg` is needed for anything but SVG) |
| `shiny` | `dbdt_explorer()` |
| `DT` | the row preview in `dbdt_explorer()` |
| `odbc` (or another DBI driver) | querying [MS SQL Server](duckdt-mssql.qmd) |

`dbdt_erd()` needs none of them.

## Attaching the package

```{r}
#| include: false
detach("package:data.table.ext", unload = TRUE)
options(duckdt.quiet = FALSE)
```

```{r}
library(data.table.ext)
```

```{r}
#| include: false
options(duckdt.quiet = TRUE)
```

The banner is the package's function index, grouped by theme: the top level is the two halves of the package, the next level is what the functions are for. [The reference](reference.qmd) has the same grouping with a line on each function.

Attaching also switches on four display enhancements by running `turn_everyone_on()`:

| What changes | What you see | Undo with |
|:---|:---|:---|
| `print()` of a `data.table` | Thousands separators in row numbers, coloured class row, an `ncol:` header, separators between groups | `disable_dt_print_thousands()` |
| `str()` of a `data.table` | A compact `# A 'data.table': 150 rows, 5 variables.` header, no `.internal.selfref` line | `disable_dt_str_mask()` |
| `dput()` of a `data.table` | No `.internal.selfref` noise, so the output can be pasted and run | `disable_dt_dput_mask()` |
| `sample_dt()`, `cdt()`, `dupe_dt()` | Colour-coded output by default | `switch_col(FALSE)` |

Call `turn_everyone_on()` at any time to put all four back.

::: {.callout-note}
## The masks live in your global environment
`print.data.table`, `str` and `dput` are defined in the global environment, so they apply to code you run at the console or in a script, and not to code inside other packages. If you have your own `print.data.table`, `str` or `dput` there, attaching replaces it, and the `disable_*()` functions remove whatever is defined.
:::

### Hiding the banner

Set the option before the package is attached, for example in your `.Rprofile`. The enhancements are still switched on; only the banner is skipped.

```r
options(duckdt.quiet = TRUE)
library(data.table.ext)
```

`suppressPackageStartupMessages(library(data.table.ext))` works too, because the whole banner is one startup message.

## First steps

### Display

```{r}
DT <- as.data.table(iris)[c(1:5, 51:55, 101:105)]
sample_dt(DT, n = 2, group = Species)
```

```{r}
str(DT)
```

More in [Printing & sampling](display.qmd) and [Data checks & helpers](checks.qmd).

### DuckDB

```{r}
con <- dbdt_example()
d <- dbdt(con, "orders")
d[customer_id == 1]
```

```{r}
#| include: false
dbdt_disconnect(con)
```

`d[i, j, by]` ran as one SQL query and returned a `data.table`. Start with [Connect & load](duckdt-connect.qmd).
