---
title: "data.table.ext"
subtitle: "Friendlier `data.table` output, and `data.table` syntax for DuckDB"
toc: false
---

```{r}
#| include: false
source("_setup.R")
```

`data.table.ext` bundles two families of `data.table` extensions in one package.

::: {.grid}

::: {.g-col-12 .g-col-md-6}
### Display, sampling & readability

Cleaner printing, presentation-friendly grouped sampling, and quick checks for the tables you explore every day.

- Thousands separators, coloured class tokens and a column count on every print
- `sample_dt()` and `cdt()`: colour-coded, group-aware sampling and display
- `highlight_dt()`, `dupe_dt()`, `outlier_dt()`: surface the rows that matter
- `schema_dt()`, `na_dt()`, `freq_dt()`, `key_dt()`, `spark_dt()`: profile a table
- Friendlier `str()` and `dput()`; markdown export with `md_dt()`

[Printing & sampling](display.qmd){.btn .btn-primary}
[Data checks & helpers](checks.qmd){.btn .btn-outline-primary}
:::

::: {.g-col-12 .g-col-md-6}
### dbdt: databases, data.table style

Write `d[i, j, by]`. It runs as SQL inside [DuckDB](https://duckdb.org/), and only the answer comes back to R as a `data.table`.

- Query in-memory data, `.duckdb` files, CSV and Parquet without loading them into R
- Joins, upserts and temporary subsets that stay in the database
- Reverse-engineer a database into a data model and draw it as an interactive ER diagram
- Pick tables and columns in the diagram; it writes the query for you

[Connect & load](duckdt-connect.qmd){.btn .btn-primary}
[Query](duckdt-query.qmd){.btn .btn-outline-primary}
:::

:::

## Install

```r
# From GitHub
devtools::install_github("KyleHaynes/data.table.ext")

# Or from a local checkout
install.packages(".", repos = NULL, type = "source")
```

`data.table`, `cli`, `DBI` and `duckdb` are installed with it. A few features use optional packages; see [Get started](getting-started.qmd#optional-packages).

## A 60-second tour

Attaching the package turns on the display enhancements. Here is grouped sampling on five rows of each `iris` species: every row for two randomly chosen species, each value coloured so repeats stand out.

```{r}
DT <- as.data.table(iris)[c(1:5, 51:55, 101:105)]
sample_dt(DT, n = 2, group = Species)
```

`dbdt_example()` builds a small in-memory DuckDB database to try things on. `dbdt()` wraps one of its tables, and from there it is ordinary `data.table` syntax:

```{r}
con <- dbdt_example()
orders <- dbdt(con, "orders")
orders[, .(n_orders = .N, first_order = min(ordered_on)), by = customer_id]
```

Draw the whole database, and the joins between its tables, in your browser:

```r
dbdt_erd(con)
```

```{r}
#| include: false
dbdt_disconnect(con)
```

## Where next

| If you want to... | Read |
|:---|:---|
| See what attaching the package changes, and how to undo it | [Get started](getting-started.qmd) |
| Make printing, sampling and `str()` output easier to scan | [Printing & sampling](display.qmd) |
| Profile a table and find duplicates, outliers and keys | [Data checks & helpers](checks.qmd) |
| Query a database file, CSV or Parquet with `data.table` syntax | [Connect & load](duckdt-connect.qmd), [Query](duckdt-query.qmd) |
| Join, update and merge inside the database | [Join, write & merge](duckdt-join-write.qmd) |
| Understand and draw a database | [Explore a database](duckdt-explore.qmd) |
| Find a function | [Function reference](reference.qmd) |
| Know what does not work yet | [Limitations](limitations.qmd) |
